Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2022-03-15 doc/peakcharacterization_manuscript.Rmd
updated for latest version of the manuscript

@@

```{r get_setup, include = FALSE}
source("./packages.R")
lapply(list.files("./R", full.names = TRUE), source)

@@

To find intervals that contain points that are not just noise, the number of non-zero intensity frequency points in each sliding window are counted.
Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. 
Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. 
However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points). 
The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, as shown in Figure XB and XC. 
The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in Figure XC and XD. 
Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE. 
This model seems to fit the known relationship between frequency and M/Z, where the frequency is related to $1/\sqrt{mz}$. 
This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the M/Z difference at a particular M/Z based on the lasso model.
Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

@@

@@

M/Z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
While scan specific models **could** be used in the conversion of M/Z to frequency, doing so results in changes to the relative peak ordering compared to M/Z space, as shown in Figure XA. 
Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.

@@

@@

In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis. 
It is expected that noise is randomly distributed across the scans. 
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise. 
The distribution of counts in sliding windows ten points wide and offset by one point is shown in Figure X. 
The number of peaks detected without applying a cutoff explodes by ..., with the vast majority of the peaks being present in a single scan. 
By discarding those windows that have a non-zero point density below the 99th percentile, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.

```{r increment_slidingwindow}
figure_count$increment("slidingwindow_count")
```

>>>>

```{r get_setup, include = FALSE}
source(here::here("packages.R"))
lapply(list.files(here::here("./R"), full.names = TRUE), source)

@@

@@

To find intervals that contain points that represent actual signal and not just random noise, the number of non-zero intensity points in each sliding window are counted.
Subsequently, we break these counts into fixed width tiles (default width of 2000) and calculate the 99th percentile of non-zero points for each tile.
The median value x 1.5 of these 99th percentile values from the fixed width tiles is used as the cutoff value to determine which of the initial sliding regions should be kept as regions that contain potential signal.
Any sliding window with a non-zero count less than or equal to the cutoff value is removed, and the remaining sliding windows are reduced, where overlapping windows are merged together to create the initial peak regions.

@@

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. 
Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. 
However, the frequency can be calculated by dividing the average M/Z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points). 
The subsequent differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, as shown in `r figure_count$label_text("mz_frequency_conversion")`B and C. 
The Thermo Fusion instrument from which most of our collaborators data has been acquired, at a resolution of 400K, has a mode of 0.5, as shown in `r figure_count$label_text("mz_frequency_conversion")`C and D. 
Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in `r figure_count$label_text("mz_frequency_conversion")`E. 
This model seems to fit the known relationship between frequency and M/Z, where the frequency is related to $1/\sqrt{mz}$. 
This is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points. 
M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model; however, it is very difficult to create a lasso model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the M/Z difference at a particular M/Z based on the lasso model.
Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

@@

@@

M/Z to frequency models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
While scan specific models **could** be used in the conversion of M/Z to frequency, doing so results in changes to the relative peak ordering compared to M/Z space, as shown in `r figure_count$label_text("peak_ordering")`. 
Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.

@@

```{r increment_slidingwindow}
figure_count$increment("slidingwindow_count")
tar_load(nonoise_vs_noise)
nn_n_ratio = english::as.english(round(nonoise_vs_noise["noperc"] / nonoise_vs_noise["perc99"]))
```

In a dataset of this nature, we expect that much of the data is really just noise, and doesn't contribute that much to the analysis. 
It is expected that noise is randomly distributed across the scans. 
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (2000 in this work), and examine the 99th percentile of the sliding window counts within the tiled region.
A histogram of sliding window counts in a single tiled region is shown in `r figure_count$label_text("slidingwindow_count")`A.
The distribution of 99th percentiles across all of the tiled regions for a single sample is shown in `r figure_count$label_text("slidingwindow_count")`B.
We can use the median (50th percentile) value of all the 99th percentiles from all of the tiled regions to determine a cutoff value above which we think any non-zero points are potentially **signal** and not just noise.
In the example below, the median of the tiled windows is 5, resulting in a cutoff of 7.5.
The number of peaks detected without applying a cutoff explodes `r nn_n_ratio`-fold (from `r format(nonoise_vs_noise["perc99"], big.mark = ",")` to `r format(nonoise_vs_noise["noperc"], big.mark = ",")` in the 97lipid sample), with a vast majority of the peaks being present in a single scan. 
By discarding those windows that have a non-zero point density below 1.5x the median of 99th percentiles, and merging the remaining adjacent windows, the regions most likely to contain actual, informative peaks are rapidly and easily determined.
